---
title: "Imputation for Meta-Analysis"
author: "Jacek Buczny"
date: "Jan 29, 2025"
output:
  pdf_document: default
  word_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)
```

```{r loading packages, include = FALSE}

library(readxl)
library(metafor)
library(writexl)
library(mice)
library(lattice)
library(gridExtra)
library(dplyr)
library(knitr)
```

## Cleaning the environment to ensure a clean run

```{r cleaning the enviroment}

rm(list=ls())
```

## Preparing to load the data

```{r listing files, include = FALSE}

list.files()

## Ensure that the correct file is in the folder
## is on the list of file. If not go to Session > Set Working Directory
## > To Source File Location
## Reading the imported data set
```

```{r loading data}

data_full <- readxl::read_xlsx("SE_SRM_missing.xlsx"
                               , sheet = "full_data_set")

data_missing <- readxl::read_xlsx("SE_SRM_missing.xlsx"
                                  , sheet = "to_impute")
```

## Meta-analysis on full data

```{r meta-analysis full}

res1_target <- rma(yi = target_effect
                   , sei = SE_target
                   , data = data_full
                   , slab = study
                   , method = "REML")

print(res1_target
      , digits = 3)

res1_perceiver <- rma(yi = perceiver_effect
                      , sei = SE_perceiver
                      , data = data_full
                      , slab = study
                      , method = "REML")

print(res1_perceiver
      , digits = 3)

res1_relationship <- rma(yi = relationship_effect
                         , sei = SE_relationship
                         , data = data_full
                         , slab = study
                         , method = "REML")

print(res1_relationship
      , digits = 3)
```

## Meta-analysis with deletions

```{r meta-analysis listwise deletion}

res2_target <- rma(yi = target_effect
                   , sei = SE_target
                   , data = data_missing
                   , slab = study
                   , method = "REML")

print(res2_target
      , digits = 3)

res2_perceiver <- rma(yi = perceiver_effect
                      , sei = SE_perceiver
                      , data = data_missing
                      , slab = study
                      , method = "REML")

print(res2_perceiver
      , digits = 3)

res2_relationship <- rma(yi = relationship_effect
                         , sei = SE_relationship
                         , data = data_missing
                         , slab = study
                         , method = "REML")

print(res2_relationship
      , digits = 3)
```

## Presenting the missing data

```{r missing table}

## Obtaining general missing information
data.frame(k.NA = colSums(is.na(data_missing)))
```

```{r}

## Obtaining the number of missing values per study
table(rowSums(is.na(data_missing)))
```

## Preparing for imputation

```{r checking variable classes}

## Ensure that all variables are numeric; use class()
class(data_missing$target_effect)
#perceiver_effect
#relationship_effect
#class(data_missing$total_N)
#SE_target			
#SE_perceiver		
#SE_relationship
```

```{r creating prediction matrix}

pred_matrix <- make.predictorMatrix(data_missing)

pred_matrix

```

### Excluding irrelevant variables

```{r prediction matrix}

## Excluding variables as predictors
pred_matrix[,"study"] <- 0
pred_matrix[,"group_N"] <- 0

## Excluding variables as predictors, BUT not preventing imputation
#pred_matrix["target_effect",] <- 0
#pred_matrix["perceiver_effect",] <- 0
#pred_matrix["relationship_effect",] <- 0

pred_matrix
```

### Selecting the imputation method

```{r imputation method}

imputation_method <- make.method(data_missing) ##PMM (ME) is the default

## Variables NOT imputed
#imputation_method["target_effect"] <- ""
#imputation_method["perceiver_effect"] <- ""
#imputation_method["relationship_effect"] <- ""

imputation_method
```

### Generating data sets with imputed values

```{r predictive mean matching}

imputed_data <- mice(data_missing
                     , print = FALSE
                     , m = 20
                     , predictorMatrix = pred_matrix
                     , method = imputation_method
                     , seed = 5678)
```

### Bayesian imputation

Bayesian imputation has been tested, but does not provide matching data.

```{r Bayesian, include = FALSE}

# imputation_method <- make.method(data_missing)
# 
# imputation_method["target_effect"] <- "norm"
# imputation_method["perceiver_effect"] <- "norm"
# imputation_method["relationship_effect"] <- "norm"
# 
# imputation_method["SE_target"] <- "norm"
# imputation_method["SE_perceiver"] <- "norm"
# imputation_method["SE_relationship"] <- "norm"
# 
# imputation_method
# 
# imputed_data <- mice(data_missing
#                      , print = FALSE, m = 20
#                      , predictorMatrix = pred_matrix
#                      , method = imputation_method
#                      , seed = 5678)

## This imputation produces negative values thus cannot be accepted.
```

```{r saving imputed data, include = FALSE}

## Saving the imputed data sets

## Looping through each imputation data set (m = 20)
## Adjust if m is different than m = 20

for (i in 1:20) {  

## Extracting each imputed data set
  imputed_dataset <- complete(imputed_data, action = i)
  
## Defining the file name
  filename <- paste0("imputed_dataset_", i, ".csv")
  
## Saving each data set as a CSV file
  write.csv(imputed_dataset, filename, row.names = FALSE)
  
## Printing message to confirm save
  print(paste("Saved:", filename))
}
```

## Validation of the selected imputation

```{r imputation summary}

## Getting basic information about imputed data
summary(imputed_data)
```

```{r plotting density}

## Extracting variables for plotting, setting labels
p1 <- densityplot(imputed_data
                  , ~ target_effect
                  , main = "Target Effect"
                  , xlab = "Effect Size")
p2 <- densityplot(imputed_data
                  , ~ perceiver_effect
                  ,  main = "Perceiver Effect"
                  , xlab = "Effect Size")
p3 <- densityplot(imputed_data
                  , ~ relationship_effect
                  , main = "Relationship Effect"
                  , xlab = "Effect Size")
p4 <- densityplot(imputed_data
                  , ~ SE_target
                  , main = "Standard Error Target Effect"
                  , xlab = "SE of the Effect Size")
p5 <- densityplot(imputed_data
                  , ~ SE_perceiver
                  , main = "Standard Error Perceiver Effect"
                  , xlab = "SE of the Effect Size")
p6 <- densityplot(imputed_data
                  , ~ SE_relationship
                  , main = "Standard Error Perceiver Effect"
                  , xlab = "SE of the Effect Size")

p1
p2
p3
p4
p5
p6

## Saving density plots as a single .png file
png("density_plots.png"
    , width = 3000
    , height = 3000
    , res = 300)

## Arranging the graphs
grid.arrange(p1
             , p2
             , p3
             , p4
             , p5
             , p6
             , ncol = 2)

dev.off()
```

```{r plotting interations}

## Presenting the interation process
plot(imputed_data)
```

### Obtaining aggregated means

```{r, include = FALSE}

## Extracting imputed data sets in long format
completed_datasets <- complete(imputed_data, "long")

## Copying source data set
final_dataset <- data_missing

## Looping through each variable with missing values
for (var in colnames(data_missing)) {
  if (any(is.na(data_missing[[var]]))) {  ## Checking if the variable has missing values
    
    ## Identifying the rows with missing values
    missing_indices <- which(is.na(final_dataset[[var]]))
    
    ## Extracting imputed values for only missing rows
    imputed_values <- completed_datasets[completed_datasets$.imp > 0, c(var, ".id")]
    
    ## Filtering only rows corresponding to missing values
    imputed_values <- imputed_values[imputed_values$.id %in% missing_indices, ]
    
    ## Computing the mean imputation for each missing observation
    mean_imputed_values <- aggregate(imputed_values[[var]], by = list(imputed_values$.id), FUN = mean)
    
    ## Ordering the mean imputed values by original row order
    mean_imputed_values <- mean_imputed_values[order(mean_imputed_values$Group.1), ]

    # Replacing only missing values in the original dataset
    final_dataset[missing_indices, var] <- mean_imputed_values$x
  }
}
```

```{r}

## Checking the final data set with mean-imputed values
head(final_dataset)

## Saving the final (aggregated) data set
write_xlsx(final_dataset, "final_data_set.xlsx")
```

### Statistical tests of the distributions

```{r, include = FALSE}

##Extracting variables
obs_target <- complete(imputed_data, action = "long")$target_effect
imp_target <- final_dataset$target_effect

obs_perceiver <- complete(imputed_data, action = "long")$perceiver_effect
imp_perceiver <- final_dataset$perceiver_effect

obs_relationship <- complete(imputed_data, action = "long")$relationship_effect
imp_relationship <- final_dataset$relationship_effect

obs_SE_target <- complete(imputed_data, action = "long")$SE_target
imp_SE_target <- final_dataset$SE_target

obs_SE_perceiver <- complete(imputed_data, action = "long")$SE_perceiver
imp_SE_perceiver <- final_dataset$SE_perceiver

obs_SE_relationship <- complete(imputed_data, action = "long")$SE_relationship
imp_SE_relationship <- final_dataset$SE_relationship

## Conduction t-tests
t.test(obs_target, imp_target)
t.test(obs_perceiver, imp_perceiver)
t.test(obs_relationship, imp_relationship)

t.test(obs_SE_target, imp_SE_target)
t.test(obs_SE_perceiver, imp_SE_perceiver)
t.test(obs_SE_relationship, imp_SE_relationship)

## Conducting Kolmogorov-Smirnov test
ks.test(obs_target, imp_target)
ks.test(obs_perceiver, imp_perceiver)
ks.test(obs_relationship, imp_relationship)

ks.test(obs_SE_target, imp_SE_target)
ks.test(obs_SE_perceiver, imp_SE_perceiver)
ks.test(obs_SE_relationship, imp_SE_relationship)
```

```{r}

## Defining a function to extract t-test results with three decimals
get_ttest_results <- function(obs, imp) {
  test <- t.test(obs, imp)
  return(c(test$statistic, test$p.value))
}

## Defining a function to extract KS-test results
get_ks_results <- function(obs, imp) {
  test <- ks.test(obs, imp)
  return(c(test$statistic, test$p.value))
}

## Performing t-tests and KS-tests for each variable
results <- data.frame(
  Variable = c("Target Effect"
               , "Perceiver Effect"
               , "Relationship Effect"
               , "SE Target"
               , "SE Perceiver"
               , "SE Relationship")
  ,
  
  T_Value = sprintf("%.3f", c(
    get_ttest_results(obs_target, imp_target)[1],
    get_ttest_results(obs_perceiver, imp_perceiver)[1],
    get_ttest_results(obs_relationship, imp_relationship)[1],
    get_ttest_results(obs_SE_target, imp_SE_target)[1],
    get_ttest_results(obs_SE_perceiver, imp_SE_perceiver)[1],
    get_ttest_results(obs_SE_relationship, imp_SE_relationship)[1]
  )),
  
  T_P_Value = sprintf("%.3f", c(
    get_ttest_results(obs_target, imp_target)[2],
    get_ttest_results(obs_perceiver, imp_perceiver)[2],
    get_ttest_results(obs_relationship, imp_relationship)[2],
    get_ttest_results(obs_SE_target, imp_SE_target)[2],
    get_ttest_results(obs_SE_perceiver, imp_SE_perceiver)[2],
    get_ttest_results(obs_SE_relationship, imp_SE_relationship)[2]
  )),
  
  KS_D_Statistic = sprintf("%.4f", c(
    get_ks_results(obs_target, imp_target)[1],
    get_ks_results(obs_perceiver, imp_perceiver)[1],
    get_ks_results(obs_relationship, imp_relationship)[1],
    get_ks_results(obs_SE_target, imp_SE_target)[1],
    get_ks_results(obs_SE_perceiver, imp_SE_perceiver)[1],
    get_ks_results(obs_SE_relationship, imp_SE_relationship)[1]
  )),
  
  KS_P_Value = sprintf("%.4f", c(
    get_ks_results(obs_target, imp_target)[2],
    get_ks_results(obs_perceiver, imp_perceiver)[2],
    get_ks_results(obs_relationship, imp_relationship)[2],
    get_ks_results(obs_SE_target, imp_SE_target)[2],
    get_ks_results(obs_SE_perceiver, imp_SE_perceiver)[2],
    get_ks_results(obs_SE_relationship, imp_SE_relationship)[2]
  ))
)

## Printing results as a properly formatted table
kable(results, caption = "Results of T-tests and Kolmogorov-Smirnov Tests")
```

## Calculating meta-analysis with imputed data

### Target effect

```{r}

res3_target <- with(imputed_data
                    , rma(yi = target_effect
                    , sei = SE_target))

## Calculating averaged coefficients
pool_target <- summary(pool(res3_target))
pool_target[-1] <- round(pool_target[-1]
                         , digits=4)

pool_target
```

### Perceiver effect

```{r}

res3_perceiver <- with(imputed_data
                       , rma(yi = perceiver_effect
                       , sei = SE_perceiver))

## Calculating averaged coefficients
pool_perceiver <- summary(pool(res3_perceiver))
pool_perceiver[-1] <- round(pool_perceiver[-1]
                            , digits=4)

pool_perceiver
```

### Relationship effect

```{r}

res3_relationship <- with(imputed_data
                          , rma(yi = relationship_effect
                          , sei = SE_relationship))

## Calculating averaged coefficients
pool_relationship <- summary(pool(res3_relationship))
pool_relationship[-1] <- round(pool_relationship[-1]
                               , digits=4)

pool_relationship
```

The following piece of code uses aggregated imputed data sets. This
indicates that the main meta-analysis can be prepared in a separate file
basd on the file *final_data_set.xlsx*.

Important: If we do not impute any of the effect sizes, studies not reporting them should be excluded. If necessary, we should write a formula for exclusion of imputed effect size and corresponding SEs.

```{r meta-analysis imputed data}

## Calculating the standard meta-analysis
res3_target <- rma(yi = target_effect
                   , sei = SE_target
                   , data = final_dataset
                   , slab = study
                   , method = "REML")

print(res3_target
      , digits = 3)

##Producing standard forest plot
metafor::forest(res3_target
                , addpred = TRUE
                , xlab = "SRM Effect: Target"
                , cex = .7
                , at = c(-0.10, 0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.70, 0.80)
                , xlim = c(-1.5, 1.5)
                , digits = c(2, 1)
                , header = "Study"
                , order = "obs") ## Ordering by the effect size values
```

## Citation of the packages

```{r citations}
citation("readxl")
citation("metafor")
citation("mice")
citation("gridExtra")
citation("writexl")
citation("dplyr")
citation("knitr")
citation("utils")
```
